---
title: "tableService: A Complete Walkthrough"
subtitle: "BDO-branded gt tables from start to finish"
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: false
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tableService)
library(gt)
library(dplyr)
```

## Overview

`tableService` is an R package for creating polished, BDO-branded gt tables. It provides:

- **Themes** — `ts_gt_base()` and `ts_gt_compressed()` for consistent styling
- **Color system** — 17 BDO color tokens and 5 palette ramps, all exported individually
- **Formatting helpers** — numeric formatting, percent change, cell coloring
- **Structural tools** — spanners, merged cells, row highlighting, horizontal rules
- **Output** — save tables to HTML or PNG with `ts_save()`

This walkthrough demonstrates every function in the package using the bundled `achv_data` dataset.

## The example data

`achv_data` contains program indicators for two operating units with results, targets, and achievement rates.

```{r}
achv_data
```

---

## 1. BDO Color System

All 17 color tokens are exported individually so you can use them directly.

```{r}
# Individual tokens
bdo_ocean_deep
bdo_jade
bdo_burgundy
```

```{r}
# Named vector of all tokens
bdo_colors
```

```{r}
# Palette list (tint ramps + diverging)
str(bdo_palettes)
```

---

## 2. Base Theme: `ts_gt_base()`

The foundation for all tables. Applies BDO slate header, charcoal body text, ocean accent border, and Trebuchet MS font.

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base()
```

### Custom header colors

Override `header_fill` with any BDO token:

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base(header_fill = bdo_ocean_deep)
```

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base(header_fill = bdo_burgundy)
```

### Text case control

The `text_case` parameter controls casing for column labels, title, row groups, and stub. Options: `"uppercase"` (default), `"lowercase"`, `"capitalize"`, or `"none"`.

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base(text_case = "capitalize") %>%
  add_labs(title = "Capitalize Example",
           subtitle = "First letter of each word is capitalized")
```

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base(text_case = "none") %>%
  add_labs(title = "Original Casing Preserved",
           subtitle = "No text transform applied")
```

---

## 3. Compressed Theme: `ts_gt_compressed()`

A tightly packed variant for dashboards and slides. Wraps `ts_gt_base()` with 1px row padding, 11px font, and compact spacing. All `ts_gt_base()` arguments pass through.

```{r}
achv_data %>%
  gt() %>%
  ts_gt_compressed()
```

```{r}
achv_data %>%
  gt() %>%
  ts_gt_compressed(
    header_fill = bdo_jade_deep,
    text_case = "capitalize"
  )
```

---

## 4. Titles and Source Notes: `add_labs()`

Quick way to add a title, subtitle, and source note.

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  add_labs(
    title = "Program Achievement Summary",
    subtitle = "FY24 Cumulative Results",
    caption = "Source: DATIM | Produced with tableService"
  )
```

---

## 5. Numeric Formatting: `format_numeric_columns()`

Automatically formats whole numbers with commas and decimals as percentages.

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  format_numeric_columns()
```

---

## 6. Column Alignment: `align_columns()`

Text columns left-aligned, numeric columns right-aligned.

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  format_numeric_columns() %>%
  align_columns()
```

---

## 7. Row Padding: `adjust_row_padding()`

Accepts keyword presets or custom pixel values.

::: {.panel-tabset}

### Condensed
```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  adjust_row_padding("condensed")
```

### Regular (default)
```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  adjust_row_padding("regular")
```

### Relaxed
```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  adjust_row_padding("relaxed")
```

### Custom (2px)
```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  adjust_row_padding(2)
```

:::

---

## 8. Row Striping: `apply_row_striping()`

Alternating row backgrounds. Defaults to `bdo_neutral_pale`; pass any color to customize.

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  apply_row_striping()
```

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  apply_row_striping(stripe_color = bdo_ocean_soft)
```

---

## 9. Darken Column Headers: `drkn_clmn_hdr()`

Override column header text color independently from the theme.

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base(header_fill = bdo_pale_charcoal) %>%
  drkn_clmn_hdr(bdo_charcoal)
```

---

## 10. Grouped Rows

Use `gt(groupname_col = ...)` with any tableService theme. Row group labels inherit the BDO styling automatically.

```{r}
achv_data %>%
  gt(groupname_col = "operating_unit") %>%
  ts_gt_base() %>%
  format_numeric_columns() %>%
  align_columns() %>%
  apply_row_striping()
```

---

## 11. Directional Arrows: `add_arrow_column_gt()`

Append a colored arrow column comparing values against a baseline or another column. Uses `bdo_jade` (up) and `bdo_red` (down) by default.

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  format_numeric_columns() %>%
  add_arrow_column_gt(Achievement, baseline = 0.90)
```

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  add_arrow_column_gt(Result, compare_to = Target,
                      up_color = bdo_ocean_deep,
                      down_color = bdo_gold)
```

---

## 12. Rate Column: `rate_column()`

Color a numeric column by pass/fail threshold.

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  format_numeric_columns() %>%
  rate_column(Achievement, threshold = 0.90)
```

---

## 13. Color Cells by Breakpoints: `ts_color_cells()`

Apply background fills to a numeric column based on breakpoints — great for heatmap-style tables. Supply `n` breakpoints and `n + 1` colors.

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  format_numeric_columns() %>%
  ts_color_cells(
    Achievement,
    breaks = c(0.5, 0.85, 1.0),
    colors = c(bdo_red_soft, bdo_gold_muted, bdo_jade_soft, bdo_ocean_soft),
    text_color = bdo_charcoal
  )
```

---

## 14. Highlight Rows: `ts_highlight_rows()`

Highlight entire rows by matching a column value or by row index.

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  format_numeric_columns() %>%
  ts_highlight_rows(
    match_col = "operating_unit",
    match_val = "Minoria",
    fill_color = bdo_ocean_soft
  )
```

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  ts_highlight_rows(rows = c(1, 6), fill_color = bdo_jade_soft, bold = TRUE)
```

---

## 15. Bold Summary Row: `ts_gt_bold_sumrow()`

Bold and border a total or summary row. Identify by row index or by matching a label.

```{r}
# Add a total row for demonstration
achv_summary <- achv_data %>%
  bind_rows(
    tibble(operating_unit = "Total", program_area = "",
           indicator = "", Result = sum(achv_data$Result),
           Target = sum(achv_data$Target),
           Achievement = sum(achv_data$Result) / sum(achv_data$Target))
  )

achv_summary %>%
  gt() %>%
  ts_gt_base() %>%
  format_numeric_columns() %>%
  ts_gt_bold_sumrow(match_col = "operating_unit", match_val = "Total",
                    fill_color = bdo_neutral_pale)
```

---

## 16. Percent Change Formatting: `ts_fmt_pct_change()`

Formats a numeric column with +/- signs and colored text. Green for gains, red for losses.

```{r}
# Create sample data with changes
change_data <- tibble(
  indicator = c("HTS_TST", "TX_NEW", "TX_CURR", "PrEP_NEW"),
  current = c(15000, 8200, 42000, 3100),
  previous = c(13500, 9000, 40000, 2800),
  pct_change = (current - previous) / previous
)

change_data %>%
  gt() %>%
  ts_gt_base() %>%
  format_numeric_columns() %>%
  ts_fmt_pct_change(pct_change)
```

---

## 17. Merge & Stack Columns: `ts_merge_stack()`

Merge two columns into one cell — primary value on top, secondary below in smaller, lighter text.

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  ts_merge_stack(indicator, program_area) %>%
  format_numeric_columns()
```

---

## 18. Themed Spanners: `ts_spanner()`

Add a column spanner that inherits BDO font styling.

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  format_numeric_columns() %>%
  ts_spanner("Performance Metrics", columns = c(Result, Target, Achievement))
```

With a custom fill color:

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  format_numeric_columns() %>%
  ts_spanner("Performance Metrics",
             columns = c(Result, Target, Achievement),
             fill_color = bdo_ocean_deep)
```

---

## 19. Horizontal Rules: `ts_hline()`

Insert visual separators between rows without using row groups.

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  format_numeric_columns() %>%
  ts_hline(rows = 7, color = bdo_ocean2, weight = 2)
```

---

## 20. Footnotes: `ts_footnote()`

Attach footnotes to column headers or specific cells.

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  format_numeric_columns() %>%
  ts_footnote("Cumulative FY24 results", columns = Result) %>%
  ts_footnote("FY24 DATIM targets", columns = Target)
```

```{r}
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  format_numeric_columns() %>%
  ts_footnote("Preliminary data", columns = Achievement, rows = c(1, 2))
```

---

## 21. Saving Tables: `ts_save()`

Save to HTML or PNG. Automatically creates the output directory.

```{r}
#| eval: false
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  format_numeric_columns() %>%
  add_labs(title = "My Report Table") %>%
  ts_save("my_table.html")

# PNG output
achv_data %>%
  gt() %>%
  ts_gt_base() %>%
  ts_save("my_table.png")
```

---

## 22. Backward Compatibility: `si_gt_base()`

`si_gt_base()` is an alias for `ts_gt_base()`. Existing code using the old name will continue to work.

```{r}
achv_data %>%
  gt() %>%
  si_gt_base() %>%
  add_labs(title = "Same as ts_gt_base()")
```

---

## Putting it all together

A kitchen-sink example combining multiple features:

```{r}
achv_data %>%
  gt(groupname_col = "operating_unit") %>%
  ts_gt_base(text_case = "capitalize") %>%
  ts_merge_stack(indicator, program_area) %>%
  format_numeric_columns() %>%
  align_columns() %>%
  apply_row_striping() %>%
  adjust_row_padding("condensed") %>%
  rate_column(Achievement, threshold = 0.90) %>%
  ts_spanner("Performance", columns = c(Result, Target, Achievement)) %>%
  ts_footnote("Cumulative FY24", columns = Result) %>%
  add_labs(
    title = "Program Achievement Dashboard",
    subtitle = "FY24 Cumulative | All Operating Units",
    caption = "Source: DATIM | Produced with tableService"
  )
```

---

## Function Reference

| Function | Purpose |
|---|---|
| **Themes** | |
| `ts_gt_base()` | Base BDO-branded theme |
| `ts_gt_compressed()` | Dense variant for dashboards |
| `si_gt_base()` | Alias for `ts_gt_base()` |
| **Formatting** | |
| `format_numeric_columns()` | Auto-format integers (commas) and decimals (%) |
| `ts_fmt_pct_change()` | +/- percent change with color |
| `align_columns()` | Text left, numbers right |
| `adjust_row_padding()` | Keywords or custom px |
| **Styling** | |
| `apply_row_striping()` | Alternating row backgrounds |
| `rate_column()` | Pass/fail color by threshold |
| `ts_color_cells()` | Heatmap fills by breakpoints |
| `ts_highlight_rows()` | Highlight rows by value or index |
| `ts_gt_bold_sumrow()` | Bold a summary/total row |
| `drkn_clmn_hdr()` | Override header text color |
| `add_arrow_column_gt()` | Directional arrows with color |
| **Structure** | |
| `ts_merge_stack()` | Stack two columns in one cell |
| `ts_spanner()` | Themed column spanner |
| `ts_hline()` | Horizontal rule between rows |
| `ts_footnote()` | Footnotes on headers or cells |
| `add_labs()` | Title, subtitle, source note |
| **Output** | |
| `ts_save()` | Save to HTML or PNG |
| **Colors** | |
| `bdo_colors` | Named vector of 17 BDO tokens |
| `bdo_palettes` | List of 5 palette ramps |
| `bdo_*` | 17 individual tokens + 5 palette vectors |
